---
title: "Project 1"
author: "Josh Stim"
date: "2023-09-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Chocolate Bars {.tabset .tabset-fade}

## Part 0. Getting started

Install relevant packages

```{r}
## Install the tidyverse package if you don't have it
if (!require("tidyverse", quietly = TRUE)) {
    install.packages("tidyverse")
}
```

Load relevant packages from library

```{r}
library(tidyverse)
library(tidytuesdayR)
library(here)
library(gapminder)
library(gghalves)
```

Upload chocolate data

```{r}
# creates data folder if one doesn't exist locally
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# saves data only if it doesn't exist locally
if (!file.exists(here("data","chocolate.RDS"))) {
  url_csv <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-18/chocolate.csv"
  chocolate <- readr::read_csv(url_csv)
  
  # save the file to RDS objects
  saveRDS(chocolate, file = here("data", "chocolate.RDS"))
}

# read in the .RDS locally from our computing environment
chocolate <- readRDS(here("data", "chocolate.RDS"))
as_tibble(chocolate)
```

## Part 1. Explore Data

### 1.1. Make a Histogram

*Make a histogram of the rating scores to visualize the overall
distribution of scores. Change the number of bins from the default to
10, 15, 20, and 25. Pick on the one that you think looks the best.
Explain what the difference is when you change the number of bins and
explain why you picked the one you did.*

```{r}
chocolate %>%
  ggplot(aes(x = rating)) +
    geom_histogram(bins = 25) +
    labs(title = "Distribution of Chocolate Bar Ratings", 
         x = "Rating", y = "Count")
```

**Josh:** Above is a histogram of the chocolate bar rating scores. I
decided to use 25 bins, since this ensures that each bin represents the
frequency of only one numerical rating. Additionally, the bars in the
histogram are centered on the rating they represent, making it easier to
interpret.

### 1.2. Count reviews for each country of bean origin

*How many reviews come from each country of bean origin?*

```{r}
n_reviews_by_cobo <- chocolate %>%
  group_by(country_of_bean_origin) %>%
  summarize(n_reviews = n()) %>%
  arrange(desc(n_reviews))

n_reviews_by_cobo
```

### 1.3. Summarize ratings of beans from Ecuador

*What is the average rating scores from reviews of chocolate bars that*
*have Ecuador as country_of_bean_origin in this dataset? For this same*
*set of reviews, also calculate*

*(1) the total number of reviews and (2) the standard deviation of the
rating scores.*

```{r}
chocolate.beans_from_equador <- 
  chocolate %>%
  filter(country_of_bean_origin == "Ecuador")

chocolate.beans_from_equador %>%
  summarize(mean = mean(rating), sd = sd(rating), total = n())
```

### 1.4. Manufacturer with best chocolate from Ecuadorian beans

*Which company manufacturer makes the best chocolate (or has the highest
ratings on average) with beans from Ecuador?*

```{r}
chocolate.beans_from_equador %>%
  group_by(company_manufacturer) %>%
  summarise(mean_rating = mean(rating)) %>%
  filter(mean_rating == max(mean_rating))
```

**Josh:** These are the five companies with the highest mean rating using beans
from Ecuador.

### 1.5. Mean ratings for all countries of origin

*Calculate the average rating across all country of origins for beans.
Which top 3 countries have the highest ratings on average?*

```{r}
mean_rating_by_bean_origin <-
  chocolate %>%
  group_by(country_of_bean_origin) %>%
  summarize(n_reviews = n(), mean_rating = mean(rating)) %>%
  arrange(desc(mean_rating))

mean_rating_by_bean_origin
```

**Josh:** We see that Tobago, China, and Sao Tome & Principe have the
highest average ratings. However, note that these countries have very
few reviews, so is it really fair to say that these countries make the
best beans?

### 1.6. Mean ratings from countries with $\geq$ 10 reviews

*Following up, remove any countries of bean origins that have less than
10 chocolate bar reviews. Now, which top 3 countries have the highest
ratings on average?*

```{r}
mean_rating_by_bean_origin %>%
  filter(n_reviews >= 10)
```

**Josh:** After removing countries with fewer than 10 reviews, we see that
**Solomon Islands**, **Congo**, and **Cuba** have the highest average
bean ratings.

### 1.7. Relationship between percent chocolate and rating

*For this last part, let's explore the relationship between percent
chocolate and ratings.*

*a) Identify the countries of bean origin with at least 50 reviews.
Remove reviews from countries are not in this list.*

**Josh:** Below is a subset of the chocolate dataset containing data from only
countries that had 50 or more reviews.

```{r}
chocolate.cobo_geq_50 <- chocolate %>%
  group_by(country_of_bean_origin) %>%
  filter(n() >= 50) %>%
  ungroup()

chocolate.cobo_geq_50
```

*(b) Using the variable describing the chocolate percentage for each
review, create a new column that groups chocolate percentages into one
of four groups: *(i)* $<60 \%$, *(ii)* $\geq 60\%$ to $<70\%$, *(iii)* $\geq 70 \%$ 
to $\leq 90 \%$, *and (iii)* $\geq 90 \%$.

*(Hint check out the substr() function in base R and the case_when()
function from dplyr -- see example below).*

```{r}
tmp <- substr(chocolate.cobo_geq_50$cocoa_percent,1,2)
chocolate.cobo_geq_50$cocoa_percent_discrete <-
  case_when(
  tmp < 60 ~ "<60%",
  tmp >= 60 & tmp < 70 ~ ">=60% to <70%",
  tmp >= 70 & tmp < 90 ~ ">=70% to <90%",
  tmp >= 90 ~ ">=90%"
  ) %>%
  factor()

levels(chocolate.cobo_geq_50$cocoa_percent_discrete)
```

*(c) Using the new column described in #2, re-order the factor levels
(if needed) to be starting with the smallest percentage group and
increasing to the largest percentage group (Hint check out the
fct_relevel() function from forcats).*

```{r}
chocolate.cobo_geq_50$cocoa_percent_discrete <- 
  chocolate.cobo_geq_50$cocoa_percent_discrete %>%
  fct_relevel(sort)

levels(chocolate.cobo_geq_50$cocoa_percent_discrete)
```

*(d) For each country, make a set of four side-by-side boxplots plotting
the groups on the x-axis and the ratings on the y-axis. These plots
should be faceted by country.*

```{r fig.width = 10, fig.height = 7}
chocolate.cobo_geq_50 %>%
  ggplot(aes(x = cocoa_percent_discrete, y = rating)) +
  geom_boxplot() +
  facet_wrap(~country_of_bean_origin) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "How Percent Cocoa affects Chocolate Ratings (by Country of Bean Origin)")
```

**Josh:** Overall, it seems that cocoa percentages between 60% and 90% yield
better ratings compared to those $<60\%$ or $>90\%$. This observation is
generally consistent across countries of bean origin, the exceptions of
Brazil and (perhaps) Blend. It is also worth noting that over a quarter
of countries produced no chocolate with $<60\%$ or $>90\%$ cocoa.

*On average, which category of chocolate percentage is most highly
rated? Do these countries mostly agree or are there disagreements?*

```{r}
#collapsing across countries
chocolate.cobo_geq_50 %>%
  group_by(cocoa_percent_discrete) %>%
  summarize(n = n(), mean = mean(rating), sd = sd(rating))
```

**Josh:** Based on the boxplots (and the table) we see that cocoa
percentages between 60% and 90% generally yield higher ratings compared
to cocoa percentages below 60% or above 90%. Additionally, it seems that
the most common cocoa percentage is $\geq 70 \%$ to $90 \%$, as nearly
84% of all chocolates with 50 or more reviews were in this category.

## Part 2. Join two datasets together

### 2.0. Load gapminder data

```{r}
gapminder.data <- unique(gapminder::gapminder[, c("country", "continent")])
```

### 2.1. Merge `gapminder$continent` to `chocolate` dataset

*Use this dataset it to create a new column called continent in our
chocolate dataset that contains the continent name for each review where
the country of bean origin is.*

*Also manually add continents that have not been matched.*

```{r}
# create copy of raw dataframe
chocolate.with_cont <- chocolate

# standardize spelling of certain countries in chocolate.with_cont.country_of_bean_origin
chocolate.with_cont$country_of_bean_origin <-
  case_when(
    chocolate.with_cont$country_of_bean_origin %in% c("Trinidad", "Tobago") ~ "Trinidad and Tobago",
    chocolate.with_cont$country_of_bean_origin %in% c("Sao Tome", "Principe", "Sao Tome & Principe") ~ "Sao Tome and Principe",
    chocolate.with_cont$country_of_bean_origin %in% c("Congo", "DR Congo") ~ "Congo, Dem. Rep.",
    chocolate.with_cont$country_of_bean_origin %in% c("U.S.A.", "USA") ~ "United States",
    .default = as.character(chocolate.with_cont$country_of_bean_origin)
  )

# join chocolate with gapminder.data by country
chocolate.with_cont <- left_join(chocolate.with_cont, gapminder.data, 
                       join_by(country_of_bean_origin == country))

# manually add continents to chocolate for countries not listed in gapminder.data
chocolate.with_cont$continent <-
  case_when(
    chocolate.with_cont$country_of_bean_origin %in% 
      c("Belize", "Grenada", "Martinique", "St. Lucia", "St.Vincent-Grenadines", "Suriname") ~ "Americas",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Burma", "Sumatra", "Sulawesi") ~ "Asia",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Ivory Coast") ~ "Africa",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Fiji", "Papua New Guinea", "Vanuatu", "Solomon Islands", "Samoa") ~ "Oceania",
    .default = as.character(chocolate.with_cont$continent)
  )
```

### 2.2 - 2.3. Remove countries named "Blend" or with \<10 reviews

*Only keep reviews that have reviews from countries of bean origin with
at least 10 reviews.* *Also, remove the country of bean origin named
"Blend".*

```{r}
chocolate.with_cont.cobo_geq_10 <- chocolate.with_cont %>%
  filter(country_of_bean_origin != "Blend") %>%
  group_by(country_of_bean_origin) %>%
  filter(n() >= 10) %>%
  ungroup()
  
chocolate.with_cont.cobo_geq_10
```

### 2.4. Make continent-by-rating violin plot

*Make a set of violin plots with ratings on the y-axis and continents on
the x-axis.*

```{r}
chocolate.with_cont.cobo_geq_10 %>%
  ggplot(aes(x = continent, y = rating)) +
  geom_violin(fill = "skyblue") +
  geom_jitter(size = 0.5, width = 0.1, alpha = 0.2) +
  stat_summary(fun = "mean", color = "white", size = 0.3) +
  theme(legend.position = "none") +
  theme_bw()
```

**Josh:** Overall, there does not seem to be a difference in mean ratings across
continents. However you may notice that a large proportion of ratings
are coming from `Americas`. This may be due to the fact that `Americas`
actually includes two continents: North and South America.

**Josh:** Also, you may notice that, while the mean ratings are similar across
continents, most of the lower rated chocolate bars (rating $\leq 2$)
come from the Americas. This may be due to the fact that we have so many
samples from `Americas`, and so it shouldn't be too surprising to see a
greater number of extreme ratings compared to the other, less sampled
continents.

## Part 3. Convert Wide Data into Long Data

### 3.1. Create an indicator column for each ingredient

*Create a new set of columns titled beans, sugar, cocoa_butter, vanilla,
letchin, and salt that contain a 1 or 0 representing whether or not that
review for the chocolate bar contained that ingredient (1) or not (0).*

```{r}
# convert missing ingredient cell to empty string
chocolate$ingredients[is.na(chocolate$ingredients)] <- " "

# create ingredient columns
chocolate.with_ingredients <- 
  chocolate %>%
  mutate(
    beans = str_detect(ingredients, "B"),
    cocoa_butter = str_detect(ingredients, "C"),
    vanilla = str_detect(ingredients, "V"),
    letchin = str_detect(ingredients, "L"),
    salt = str_detect(ingredients, "Sa"),
    sugar = str_detect(ingredients, "S")
  )
```

### 3.2. Create an indicator column for each characteristic

*Create a new set of columns titled char_cocoa, char_sweet, char_nutty,
char_creamy, char_roasty, char_earthy that contain a 1 or 0 representing
whether or not that the most memorable characteristic for the chocolate
bar had that word (1) or not (0). For example, if the word "sweet"
appears in the most_memorable_characteristics, then record a 1,
otherwise a 0 for that review in the char_sweet column (Hint: check out
str_detect() from the stringr package).*

```{r}
chocolate.with_ingredients <- 
  chocolate.with_ingredients %>%
  mutate(
    char_cocoa = str_detect(most_memorable_characteristics, "cocoa"),
    char_sweet = str_detect(most_memorable_characteristics, "sweet"),
    char_nutty = str_detect(most_memorable_characteristics, "nutty"),
    char_creamy = str_detect(most_memorable_characteristics, "creamy"),
    char_roasty = str_detect(most_memorable_characteristics, "roasty"),
    char_earthy = str_detect(most_memorable_characteristics, "earthy")
  )
```

### 3.3. Percent of bars that had each ingredient/characteristic (by year)

*For each year (i.e. review_date), calculate the mean value in each new
column you created across all reviews for that year. (Hint: If all has
gone well thus far, you should have a dataset with 16 rows and 13
columns).*

```{r}
table.mean_ingredients_by_year <- 
  chocolate.with_ingredients %>%
  select(review_date,beans:char_earthy) %>%
  group_by(review_date) %>%
  summarize(across(beans:char_earthy, mean))

table.mean_ingredients_by_year
```

### 3.4. Convert wide dataset to long

*Convert this wide dataset into a long dataset with a new feature and
mean_score column.*

```{r}
table.mean_ingredients_by_year.long <-
  pivot_longer(table.mean_ingredients_by_year,
               cols = beans:char_earthy,
               names_to = "ingredient",
               values_to = "mean_score") %>%
  as.tibble()

table.mean_ingredients_by_year.long$ingredient <-
  table.mean_ingredients_by_year.long$ingredient %>%
  factor() %>%
  fct_relevel(
    c("beans", "cocoa_butter","letchin", "salt", "sugar", "vanilla",
      "char_cocoa", "char_creamy", "char_earthy", "char_nutty", "char_roasty", "char_sweet")
    )

table.mean_ingredients_by_year.long
```

## Part 4. Data Visualization

### 4.1 - 4.4. Visualize changes in chocolate composition over the years

*Use the functions in ggplot2 package to make a scatter plot of the
mean_scores (y-axis) over time (x-axis). One point for each mean_score.
For full credit, your plot should include:*

*1) An overall title for the plot and a subtitle summarizing key trends
that you found. Also include a caption in the figure with your name.*

*2) Both the observed points for the mean_score, but also a smoothed
non-linear pattern of the trend*

*3) All plots should be shown in the one figure*

*4) There should be an informative x-axis and y-axis label*

```{r}
table.mean_ingredients_by_year.long %>%
  ggplot(aes(x = review_date, y = mean_score, color = ingredient)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    se = FALSE
  ) +
  facet_wrap(~ingredient) +
  theme(legend.position = "none") +
  theme_bw() +
  labs(title = "How have chocolate ingredients and characteristics change over time?",
       subtitle = "Based on review data from 2006 to 2021",
       x = "Year",
       y = "Proportion of bars with feature")
```

**Josh:** Overall, it seems that over the years, less letchin and vanilla are
being included in chocolate bars. Cocoa butter has seen as small dip in
popularity, but not as pronounced as that of letchin and vanilla. Beans
and sugar are, by far, the most popular chocolate ingredient, with
nearly every bar including them in their ingredient profile.

## Part 5. Make the worst plot you can!

### 5.1 - 5.2. Create a terrible plot

*Using the chocolate dataset (or any of the modified versions you made
throughout this assignment or anything else you wish you build upon
it):*

*Make the absolute worst plot that you can. You need to customize it in
at least 7 ways to make it awful. In your document, write 1 - 2
sentences about each different customization you added (using bullets --
i.e. there should be at least 7 bullet points each with 1-2 sentences),
and how it could be useful for you when you want to make an awesome data
visualization.*

**Josh:** My terrible plot is shown below.

```{r}
chocolate.with_cont.cobo_geq_10 %>%
  ggplot(aes(x = factor(country_of_bean_origin), fill = factor(review_date), y = rating)) +
  geom_boxplot() +
  facet_wrap(~review_date) +
  labs(title = "Rating by review year and country of bean origin") +
  theme(axis.title.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        panel.grid.major = element_line(color = "blue"),
        plot.background = element_rect(fill = "red"),
        plot.title = element_text(size = 25))
```

**Josh:** Here are the customizations I made.

1)  `aes(x = factor(country_of_bean_origin), fill = factor(review_date), y = rating)`:
    I created a boxplot with `country_of_bean_origin` on the x-axis,
    `rating` on the y-axis, and `review_date` as the fill variable.
    There are 34 countries, and 16 review dates, so the plot will
    contain 34 \* 16 = 544 boxes (assuming there is data for every
    country-year combo). This will be useful for making a great data visualization 
    because it will allow me to easily toggle how I define my x, y, and fill variables.

2)  `geom_boxplot()` This customization created a boxplot from my ggplot
    object. I think this method of creating a boxplot is particularly
    useful because it is so easy to customize.

3)  `facet_wrap(~review_date)` I used this to create 16 facets, each
    defined by a specific year between 2006 and 2021. This customization
    is useful for separating visualizations by groups and makes it
    easier to see relationships within groups.

4)  `axis.title.y = element_text(size = 5)` I made the y-axis label have
    a font-size of 5. This makes it very small and therefore difficult
    to read. A better plot would have a larger font size.

5)  `axis.title.x = element_text(size = 5)` I also made the x-axis label
    have a font-size of 5. Again, a better plot would have a larger font
    size.

6)  `panel.grid.major = element_line(color = "blue")` This changed the
    color of the grid lines on each facet. Not really necessary for this
    plot, but could be necessary if the analyst wants to make it easier
    to read coordinates on the plot.

7)  `plot.background = element_rect(fill = "red")` Makes the canvas red.
    This is completely unnecessary for this plot, and makes it look
    quite ugly. I suppose this specification could be useful if someone
    wants to adjust the contrast between their plot and the canvas.

8)  `plot.title = element_text(size = 25)` I made the title have a
    font-size of 25. This is simply too large for this already
    overloaded plot. A better visualization would use a smaller font
    size for the title.
    
## Part 6. Make my plot better!

### 6.1 - 6.2. Improve a plot

*The goal is to take my sad looking plot and make it better!*

*You need to customize it in at least 7 ways to make it better. In your
document, write 1 - 2 sentences about each different customization you
added (using bullets -- i.e. there should be at least 7 bullet points
each with 1-2 sentences), describing how you improved it.*

**Josh:** I've reposted the original plot for reference:

```{r}
chocolate %>%
    ggplot(aes(
        x = as.factor(review_date),
        y = rating,
        fill = review_date
    )) +
    geom_violin()
```

**Josh:** And here is my new-and-improved plot.

```{r fig.width = 10, fig.height = 7}
chocolate %>%
  ggplot(aes(
      x = as.factor(review_date),
      y = rating
    )) +
  geom_half_violin(position = position_nudge(x = -0.15), fill = "chocolate") +
  geom_jitter(size = 0.5, width = 0.075, alpha = 0.35) +
  labs(title = "Chocolate ratings become less variable over time",
       subtitle = "A visualization of chocolate ratings by year of review (from 2006 to 2021)",
       caption = "Created by Josh Stim using `gghalves` and `ggplot2`") +
  xlab("Year of Review") +
  ylab("Rating") +
  theme_bw(base_family = "Times", base_size = 15) +
  theme(axis.title.x = element_text(size = 18, vjust = -2),
        axis.title.y = element_text(size = 18, vjust = 2),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 9, vjust = -5),
        plot.title = element_text(size = 20, vjust = 2),
        plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "inches"))
```

**Josh:** I made the following customizations:

1)  `geom_half_violin(position = position_nudge(x = -0.15), fill = "chocolate", scale = "area", trim = TRUE)`
    Violin plots are redundant, in that they show the same density
    twice. I decided to use a half-violin plot (from library
    `gghalves`), which is exactly what it sounds like: it allows you to
    plot only half of the violin. I also nudged each half-violin 0.15
    x-units to the left and colored all of them the same color (having
    different colors is unnecessary and might cause confusion).

2)  `geom_jitter(size = 0.5, width = 0.075, alpha = 0.35)` I also wanted
    to plot the individual data points within each year, so the viewer
    could get a better idea of how many points were within each year. I
    adjusted the size of the points, the width of the jitter, and made
    the points translucent by setting `alpha = 0.35`.

3)  `labs(title = "Chocolate ratings become less variable over time",  subtitle = "A visualization of chocolate ratings by year of review (from 2006 to 2021)",  caption = "Created by Josh Stim using the gghalves and ggplot2")`
    Created an informative title and subtitle for the plot. Also created
    a caption to document who created the visualization, and with what R
    packages.

4)  `theme_bw(base_family = "Times", base_size = 15)` I really like the
    `theme_bw()` ggplot theme. I think it looks very clean, and provides
    a boundary that makes it easy to visually separate the plot from the
    canvas. I also changed the font to `"Times"` and `base_size` to 15
    to make the axis ticks pop out more.

5)  `axis.title.x = element_text(size = 18, vjust = -2)` and
    `axis.title.y = element_text(size = 18, vjust = 2)` Adjusted the
    size of the axis titles to make them easier to read. Also moved them
    further away from the plot in efforts to make the plot feel less
    cramped.

6)  `axis.text.x = element_text(size = 12)` and
    `axis.text.y = element_text(size = 12)` 
    Made the axis tick labels
    larger so they are easier to read. I think that this is important so
    that the viewer can easily understand the coordinates in the plot.

7)  `plot.title = element_text(size = 20, hjust = 0.5, vjust = 2)` Made
    the plot title a bit larger and centered it. Also increased the
    distance between the plot and the title, again in efforts to make
    the plot look less cramped.

8)  `plot.caption = element_text(size = 9, vjust = -5)` Reduced the size
    of the caption so it doesn't interfere with the axis title. Also
    moved the caption down on the canvas for the same reason.

9)  `plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "inches")` Increased the
    plot margins so that it takes up a greater proportion of the canvas.
    This can be useful in improving the readability of the graph.

## Part 7. R session info

```{r}
options(width = 120)
sessionInfo()
```
