---
title: "Project 1: Chocolate Bars"
author: "Josh Stim"
date: "2023-09-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Part 0: Getting started

Install relevant packages

```{r}
install.packages("tidyverse")
install.packages("tidytuesdayR")
```

Load relevant packages from library

```{r}
library(tidyverse)
library(tidytuesdayR)
library(here)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forcats)
library(gapminder)
```

Upload chocolate data

```{r}
# creates data folder if one doesn't exist locally
if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

# saves data only if it doesn't exist locally
if (!file.exists(here("data","chocolate.RDS"))) {
  url_csv <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-18/chocolate.csv"
  chocolate <- readr::read_csv(url_csv)
  
  # save the file to RDS objects
  saveRDS(chocolate, file = here("data", "chocolate.RDS"))
}

# read in the .RDS locally from our computing environment
chocolate <- readRDS(here("data", "chocolate.RDS"))
as_tibble(chocolate)
```

Part 1: Explore Data

1)  Below, I created a histogram of the chocolate bar rating scores. I
    decided to use 25 bins, since this ensures that each bin represents
    the frequency of only one numerical rating. Additionally, the bars
    in the histogram are centered on the rating they represent, making
    it easier to interpret.

```{r}
ratings.hist.bin13 <- ggplot(data = chocolate, aes(x = rating)) +
  geom_histogram(bins = 13) +
  labs(title = "Distribution of Chocolate Bar Ratings", 
       x = "Rating", y = "Count")

ratings.hist.bin13

ratings.hist.bin25 <- ggplot(data = chocolate, aes(x = rating)) +
  geom_histogram(bins = 25) +
  labs(title = "Distribution of Chocolate Bar Ratings", 
       x = "Rating", y = "Count")

ratings.hist.bin25
```

2)  Consider the countries where the beans originated from. How many
    reviews come from each country of bean origin?

```{r}
n_reviews_by_cobo <- chocolate %>%
  group_by(country_of_bean_origin) %>%
  summarize(n_reviews = n()) %>%
  arrange(desc(n_reviews))
```

3)  What is average rating scores from reviews of chocolate bars that
    have Ecuador as country_of_bean_origin in this dataset? For this
    same set of reviews, also calculate
    
(1) the total number of reviews and (2) the standard deviation of the
    rating scores. Your answer should be a new data frame with these
    three summary statistics in three columns. Label the name of these
    columns mean, sd, and total.

```{r}
chocolate.beans_from_equador <- chocolate %>%
  filter(country_of_bean_origin == "Ecuador")

chocolate.beans_from_equador %>%
  summarize(mean = mean(rating), sd = sd(rating), total = n())

chocolate.beans_from_equador
```

4)  Which country makes the best chocolate (or has the highest ratings
    on average) with beans from Ecuador?

```{r}
chocolate.beans_from_equador %>%
  group_by(company_location) %>%
  summarise(mean_rating = mean(rating)) %>%
  filter(mean_rating == max(mean_rating))
```

5)  Calculate the average rating across all country of origins for
    beans. Which top 3 countries have the highest ratings on average?

```{r}
mean_rating_by_bean_origin <- chocolate %>%
  group_by(country_of_bean_origin) %>%
  summarize(n_reviews = n(), mean_rating = mean(rating)) %>%
  arrange(desc(mean_rating))

mean_rating_by_bean_origin
```

Following up on the previous problem, now remove any countries of bean
origins that have less than 10 chocolate bar reviews. Now, which top 3
countries have the highest ratings on average?

```{r}
mean_rating_by_bean_origin %>%
  filter(n_reviews >= 10)
```

7)  For this last part, let's explore the relationship between percent
    chocolate and ratings.

a)  Identify the countries of bean origin with at least 50 reviews.
    Remove reviews from countries are not in this list.

```{r}
chocolate.cobo_geq_50 <- chocolate %>%
  group_by(country_of_bean_origin) %>%
  filter(n() >= 50) %>%
  ungroup()

chocolate.cobo_geq_50
```

(b) Using the variable describing the chocolate percentage for each
    review, create a new column that groups chocolate percentages into
    one of four groups: (i) \<60%, (ii) \>=60 to \<70%, (iii) \>=70 to
    \<90%, and (iii) \>=90% (Hint check out the substr() function in
    base R and the case_when() function from dplyr -- see example
    below).

```{r}
tmp <- substr(chocolate.cobo_geq_50$cocoa_percent,1,2)
chocolate.cobo_geq_50$cocoa_percent_discrete <-
  case_when(
  tmp < 60 ~ "<60%",
  tmp >= 60 & tmp < 70 ~ ">=60% to <70%",
  tmp >= 70 & tmp < 90 ~ ">=70% to <90%",
  tmp >= 90 ~ ">=90%"
  )
```

(c) Using the new column described in #2, re-order the factor levels (if
    needed) to be starting with the smallest percentage group and
    increasing to the largest percentage group (Hint check out the
    fct_relevel() function from forcats).

```{r}
chocolate.cobo_geq_50$cocoa_percent_discrete <- 
  chocolate.cobo_geq_50$cocoa_percent_discrete %>%
  factor() %>%
  fct_relevel(sort)
```

(d) For each country, make a set of four side-by-side boxplots plotting
    the groups on the x-axis and the ratings on the y-axis. These plots
    should be faceted by country.

```{r}
boxplot.rating_by_country <- 
  ggplot(data = chocolate.cobo_geq_50, aes(x = cocoa_percent_discrete, y = rating)) +
  geom_boxplot() +
  facet_wrap(~country_of_bean_origin)

boxplot.rating_by_country
```

```{r}
#collapsing across countries
chocolate.cobo_geq_50 %>%
  group_by(cocoa_percent_discrete) %>%
  summarize(n = n(), mean = mean(rating), sd = sd(rating))

#means within countries
chocolate.cobo_geq_50 %>%
  group_by(country_of_bean_origin, cocoa_percent_discrete) %>%
  summarize(n = n(), mean = mean(rating), sd = sd(rating)) %>%
  arrange(cocoa_percent_discrete)
```
Based on the boxplot and summary table, it seems like the cocoa
percentages with highest ratings are between 60% and 90%. Overall, this
trend is observed in all countries that produced cocoa beans.

Part 2: Join two datasets together

Load gapminder data
```{r}
gapminder.data <- unique(gapminder::gapminder[, c("country", "continent")])
```

Task 1) Make a new column called `continent` in `chocolate` dataset that assigns a continent to each country_of_bean_origin.
Also manually add continents that have not been matched.
```{r}
# create copy of raw dataframe
chocolate.with_cont <- chocolate

# standardize spelling of certain countries in chocolate.with_cont.country_of_bean_origin
chocolate.with_cont$country_of_bean_origin <-
  case_when(
    chocolate.with_cont$country_of_bean_origin %in% c("Trinidad", "Tobago") ~ "Trinidad and Tobago",
    chocolate.with_cont$country_of_bean_origin %in% c("Sao Tome", "Principe", "Sao Tome & Principe") ~ "Sao Tome and Principe",
    chocolate.with_cont$country_of_bean_origin %in% c("Congo", "DR Congo") ~ "Congo, Dem. Rep.",
    chocolate.with_cont$country_of_bean_origin %in% c("U.S.A.", "USA") ~ "United States",
    .default = as.character(chocolate.with_cont$country_of_bean_origin)
  )

# join chocolate with gapminder.data by country
chocolate.with_cont <- left_join(chocolate.with_cont, gapminder.data, 
                       join_by(country_of_bean_origin == country))

# manually add continents to chocolate for countries not listed in gapminder.data
chocolate.with_cont$continent <-
  case_when(
    chocolate.with_cont$country_of_bean_origin %in% 
      c("Belize", "Grenada", "Martinique", "St. Lucia", "St.Vincent-Grenadines", "Suriname") ~ "Americas",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Burma", "Sumatra", "Sulawesi") ~ "Asia",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Ivory Coast") ~ "Africa",
    chocolate.with_cont$country_of_bean_origin %in%
      c("Fiji", "Papua New Guinea", "Vanuatu", "Solomon Islands", "Samoa") ~ "Oceania",
    .default = as.character(chocolate.with_cont$continent)
  )

```

Task 2 & 3) Remove country of bean origin named "Blend" and remove countries with fewer than 10 reviews
```{r}
chocolate.with_cont.cobo_geq_10 <- chocolate.with_cont %>%
  filter(country_of_bean_origin != "Blend") %>%
  group_by(country_of_bean_origin) %>%
  filter(n() >= 10) %>%
  ungroup()
  
chocolate.with_cont.cobo_geq_10
```

Task 4) Make violin plots with ratings on y-axis and continent on x-axis
```{r}
violin.ratings_by_continent <-
  ggplot(data = chocolate.with_cont.cobo_geq_10, aes(x = continent, y = rating, fill = continent)) +
  geom_violin() +
  geom_jitter(width = 0.05) +
  stat_summary(fun = "mean", color = "red", size = 0.3) +
  theme(legend.position = "none")

violin.ratings_by_continent
```
Part 3: Convert Wide Data into Long Data

1) Create a new set of columns titled beans, sugar, cocoa_butter, vanilla, letchin, and salt that contain a 1 or 0 representing whether or not that review for the chocolate bar contained that ingredient (1) or not (0).

```{r}
# convert missing ingredient cell to empty string
chocolate$ingredients[is.na(chocolate$ingredients)] <- " "

# create ingredient columns
chocolate.with_ingredients <- 
  chocolate %>%
  mutate(
    beans = str_detect(ingredients, "B"),
    cocoa_butter = str_detect(ingredients, "C"),
    vanilla = str_detect(ingredients, "V"),
    letchin = str_detect(ingredients, "L"),
    salt = str_detect(ingredients, "Sa"),
    sugar = str_detect(ingredients, "S")
  )
```

2) Create a new set of columns titled char_cocoa, char_sweet, char_nutty, char_creamy, char_roasty, char_earthy that contain a 1 or 0 representing whether or not that the most memorable characteristic for the chocolate bar had that word (1) or not (0). For example, if the word “sweet” appears in the most_memorable_characteristics, then record a 1, otherwise a 0 for that review in the char_sweet column (Hint: check out str_detect() from the stringr package).

```{r}
chocolate.with_ingredients <- 
  chocolate.with_ingredients %>%
  mutate(
    char_cocoa = str_detect(most_memorable_characteristics, "cocoa"),
    char_sweet = str_detect(most_memorable_characteristics, "sweet"),
    char_nutty = str_detect(most_memorable_characteristics, "nutty"),
    char_creamy = str_detect(most_memorable_characteristics, "creamy"),
    char_roasty = str_detect(most_memorable_characteristics, "roasty"),
    char_earthy = str_detect(most_memorable_characteristics, "earthy")
  )
```

3) For each year (i.e. review_date), calculate the mean value in each new column you created across all reviews for that year. (Hint: If all has gone well thus far, you should have a dataset with 16 rows and 13 columns).

```{r}
table.mean_ingredients_by_year <- 
  chocolate.with_ingredients %>%
  select(review_date,beans:char_earthy) %>%
  group_by(review_date) %>%
  summarize(across(beans:char_earthy, mean))

table.mean_ingredients_by_year
```

4) Convert this wide dataset into a long dataset with a new feature and mean_score column.

```{r}
table.mean_ingredients_by_year.long <-
  pivot_longer(table.mean_ingredients_by_year,
               cols = beans:char_earthy,
               names_to = "ingredient",
               values_to = "mean_score")

table.mean_ingredients_by_year.long
```

Part 4: Data Visualization

```{r}
scatter.ingredient_score_over_time <- 
  ggplot(data = table.mean_ingredients_by_year.long, 
         aes(x = review_date, y = mean_score, color = ingredient)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    se = FALSE
  ) +
  facet_wrap(~ingredient) +
  theme(legend.position = "none")

scatter.ingredient_score_over_time
```